### 2013-07-17 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
cleaning up nonlinear chapter

Isocurves of the velocity magnitudes are given in
Figure~\ref{fig:driven_velocity_isocurves} for solutions to the lid
driven cavity problem at Reynolds numbers of 100, 300, 500 and
700. Increasing the velocity on the top boundary of the problem drives
up the Reynolds number of the system making it more difficult to
solve. More rotation of the fluid is induced by intertial forces and
we begin to see some additional vortices form besides the primary
vortex in the center of the box.

@@

Given the difficulty of these benchmark problems, the linear solve at
each Newton step was preconditioned with an algebraic multigrid
method, a common technique for preconditioning these equations
\citep{ghia_high-re_1982,evans_enhanced_2007}. For FANM, this
preconditioning was required to reduce the spectral radius of the
Jacobian operators generated at each Newton step and for the
Newton-Krylov solutions this preconditioning was required to obtain
good convergence properties although not necessary for
convergence. For every benchmark calculation, both FANM and
Newton-Krylov were preconditioned with the same multigrid parameters
such that the conditioning of the system would not be a factor in the
verification of the method or the perfomance comparison in the
following section.

As with the radiation transport calculations in the previous chapter,
algrebraic preconditioning with the multigrid method results in dense
linear systems. To mitigate this, the reduced domain approximation was
again applied to the Monte Carlo solution within the MCSA solve
occuring at each nonlinear iteration. The values used for the
approximation will be given for each benchmark problem. Without
multigrid preconditioning or too small of a reduced domain fill level,
the residual in MCSA was observed to stagnate. Typically this happens
because the Richardson iteration will damp the higher order error
modes on the scale of the computational grid used to solve the problem
but will miss lower order error modes. In terms of Monte Carlo, this
stagnation manifests itself in histories that simply do no terminate
and therefore no solution is acquired.

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\citep{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes Newton-Krylov methods to generate a fully
consistent solution scheme for the Navier-Stokes equations among other
systems with Jacobian generation through automatic differentiation. To
achieve these results, a Monte Carlo solver generated for this work
was injected into the nonlinear toolchain within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same production implementation of GMRES from the Trilinos
scientific computing library Aztec \citep{heroux_overview_2005} used
for the verification of MCSA for the $SP_N$ solutions in the previous
chapter was also used here as a means of comparison.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and
\sn{1}{6}. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term was computed at each Newton step
using method 2 from Shadid's work. In this forcing term method,
nonlinear residuals from the two previous Newton steps are used to
inform the choice of the forcing term for the next step.

@@

For the verification calculations, at each Rayleigh number variation
the global minimum, maximum, and average values were recorded for
fluid velocity in the x and y directions, fluid pressure, and fluid
temperature. Tables~\ref{tab:convection_ra1e3_results},
\ref{tab:convection_ra1e4_results},
\ref{tab:convection_ra1e5_results}, and
\ref{tab:convection_ra1e6_results} give the results of these
computations. It should be noted that the negative pressures computed
are due to a relative pressure being reported. For each of the data
points presented, both Newton-Krylov solutions and FANM solutions
computed identical numerical results for all of these global
values. No variation in solution was noted for each of the
computations within the tolerance of the converged nonlinear residual
norms given in
Table~\ref{tab:convection_residual_norm_comparison}. Variations in the
convergence behavior and differences in performance for these
solutions will be analyzed in the following section.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Reynolds numbers
of 100, 300, 500, and
700. Table~\ref{tab:thermal_convection_parameters} gives the physical
parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Reynolds
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} fixed and varying the
fluid velocity in the $x$ direction on the top boundary. In addition
to multigrid preconditioning for the linear solution at each Newton
step, a polynomial line search method was used as a means of
globalization assisting the Newton correction computation as outlined
by Pawlowski in \citep{pawlowski_globalization_2006} while the forcing
term was again computed using method 2 from Shadid's work
\citep{shadid_inexact_1997}.

@@

In the literature on this benchmark, solutions at Reynolds numbers of
up to \sn{1}{4} were achieved using a variety of nonlinear solution
techniques. With the FANM method, at Reynolds numbers above 700
convergence of the linear iterations could not be achieved regardless
of how agressive the multigrid preconditioner was set. In these cases,
the Jacobian operators are ill-conditioned enough that even GMRES
converges slowly relative to the thermal convection cavity
benchmark. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case and others where transport driven
by intertial forces is dominant. Because of this, only solutions to
the lid driven cavity benchmark at Reynolds numbers of up to 700 are
presented.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Krylov
method. For each benchmark, we will analyze the iterative performance
of both the Newton solver and the linear solver used to compute the
correction vector at each step. In addition, timing results will be
discussed.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method. Limiting the number of nonlinear iterations required for
solution is critical for efficiency as nonlinear function evaluations,
Jacobian generation, and preconditioner construction are typically
required at every Newton iteration. For all Rayleigh numbers except
\sn{1}{5} the number of Newton iterations required to converge is the
same while the \sn{1}{5} result required an extra iteration. This is
due to the fact that the convergence of the linear residual has a
strong effect on the convergence of the nonlinear residual, much
stronger that the effect on the convergence of the eigenvalue in the
$SP_N$ computations.

As outlined in our discussion on inexact Newton methods, depending on
how converged the linear residual is at each step, it is possible to
throw of the nonlinear iteration and converge on the wrong nonlinear
solution. Although the same preconditioner was used with both the
Newton-Krylov and FANM calculations, the convergence properties of
MCSA and GMRES are different and we therefore expect them to produce
different results at each nonlinear iteration which will in turn
effect the forcing term calculation. If they are different enough, it
is possible that another Newton iteration may be required. Even
considering the extra Newton iteration, the solutions are converged to
effectivly the same residual norm tolerance and produce the same
results as provided by the
verification.

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case. We see that the extra
iteration comes from slightly slower quadratic convergence with FANM
for this particular case. The difference in convergence of the linear
problem comes from the fact that the same multigrid preconditioner
parameters are used for this case as for the \sn{1}{3} and \sn{1}{4}
Rayleigh number cases. As the Rayleigh number increases, the problem
becomes more difficult to solve and therefore a more agressive
preconditioner should maintain iterative performance of the nonlinear
method. This was observed to be the case as for the \sn{1}{6}
Rayleigh number case the number of multigrid cycle applications was
bumped up to 3 from 1 used for the other cases and FANM converged in
the same number of nonlinear iterations as the Newton-Krylov method.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear
iterations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solver preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme. Even the
\sn{1}{5} Rayleigh number case that required more Newton iterations
required 5 fewer MCSA iterations to converge than GMRES iterations.

@@

Although the iterative performance of FANM is excellent for the
thermal convection cavity problem, the explicit preconditioning
strategy causes the same issues as was observed for the $SP_N$
equations in the previous
chapter. Table~\ref{tab:convection_speedup_comparison} gives the
speedup of the Newton-Krylov method over FANM. For all Rayleigh
numbers, we see a fairly constant speedup of O(400) when using
Newton-Krylov over FANM meaning that the implementation of the FANM
method used here is significantly slower. As the inexact Newton scheme
in Drekar was modified by simply swapping out the linear solver used
at each Newton step, this increased compute time is coming directly
from using MCSA as the solver and the overhead due to
preconditioning. When preconditioning overhead for the $SP_N$
equations was considered as shown by the data in
Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
approximately two orders of magnitude was noted between the Krylov
methods and MCSA showing consistency between those results and the
fluid benchmark results. In addition, the fairly constant speedup
values show that FANM has effectively the same time complexity as a
Newton-Krylov method and the implementation timings simply differ by a
constant value.

@@

For the lid driven cavity benchmark, performance results were again
compared between the FANM and Newton-Krylov calculations. In this
case, the nonlinear iterative performance results as given by
Table~\ref{tab:driven_nonlinear_iter_comparison} show the same number
of Newton iterations were required to converge the 100, 300, and 500
Reynolds number cases while the FANM calculation with a Reynolds
number of 700 actually converged in 4 fewer iterations. The nonlinear
residual norm as a function of Newton iteration for this case is given
in Figure~\ref{fig:re700_convergence}.

@@

The difficulty of this benchmark problem as compared to the thermal
convection cavity problem is readily apparent in the linear iteration
counts as given by Table~\ref{driven_linear_iter_comparison}. As the
opposite of what was observed for that benchmark, lid driven cavity
results show that the linear models were significantly more
ill-conditioned, requiring many more MCSA and GMRES iterations. For
all cases except the 700 Reynolds number case many more MCSA
iterations were required to converge with 64\% as many GMRES
iterations required as MCSA iterations for the 100 Reynolds number
case. For the 700 Reynolds number case, fewer MCSA iterations were
observed due to the fact that 4 fewer nonlinear iterations were
performed. The average number of linear iterations per nonlinear
iterations was still much larger for the FANM calculations.

These iterative results again point out the fundamental flaw of
requiring a spectral radius of less than unity for the MCSA iteration
to converge at each Newton step. As the problem becomes more dominated
by inertial forces, the resulting linear models become increasingly
stiff and difficult to precondition. As a result, convergence requires
many more iterations that need for the flow problem driven by
convection in the previous benchmark if convergence is achieved at
all.

>>>>

Isocurves of the velocity magnitudes are given in
Figure~\ref{fig:driven_velocity_isocurves} for solutions to the lid
driven cavity problem at Reynolds numbers of 100, 300, 500 and
700. Increasing the velocity on the top boundary of the problem drives
up the Reynolds number of the system making it more difficult to
solve. More rotation of the fluid is induced by inertial forces and
we begin to see some additional vortices form besides the primary
vortex in the center of the box.

@@

Given the difficulty of these benchmark problems, the linear solve at
each Newton step was preconditioned with an algebraic multigrid
method, a common technique for preconditioning these equations
\citep{ghia_high-re_1982,evans_enhanced_2007}. Without multigrid
preconditioning, the residual in MCSA was observed to
stagnate. Typically this happens because the Richardson iteration will
damp the higher order error modes on the scale of the computational
grid used to solve the problem but will miss lower order error
modes. In terms of Monte Carlo, this stagnation manifests itself in
histories that simply do no terminate and therefore no solution is
acquired.

For FANM, this preconditioning was required to reduce the spectral
radius of the Jacobian operators generated at each Newton step and for
the Newton-Krylov solutions this preconditioning was required to
obtain good convergence properties although it was not necessary for
convergence. For every benchmark calculation, both FANM and
Newton-Krylov were preconditioned with the same multigrid parameters
such that the conditioning of the system would not be a factor in the
verification of the method or the performance comparison in the
following section. As with the radiation transport calculations in the
previous chapter, algrebraic preconditioning with the multigrid method
results in dense linear systems. To mitigate this, the reduced domain
approximation was again applied to the Monte Carlo solution within the
MCSA solve occuring at each nonlinear iteration. The values used for
the approximation will be given for each benchmark problem.

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\citep{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes Newton-Krylov methods to generate a fully
consistent solution scheme for the Navier-Stokes equations among other
systems with Jacobian generation through automatic differentiation. To
achieve these results, a Monte Carlo solver generated for this work
was injected into the nonlinear toolchain within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same production implementation of GMRES from the Trilinos
scientific computing library Aztec \citep{heroux_overview_2005} used
for the verification of MCSA for the $SP_N$ solutions in the previous
chapter was also used here as a means of comparison and FANM
benchmarking.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
square grid. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term was computed at each Newton step
using method 2 from Shadid's work. In this forcing term method,
nonlinear residuals from the two previous Newton steps are used to
inform the choice of the forcing term for the next step.

@@

For the verification calculations, at each Rayleigh number variation
the global minimum, maximum, and average values were recorded for
fluid velocity in the x and y directions, fluid pressure, and fluid
temperature. Tables~\ref{tab:convection_ra1e3_results},
\ref{tab:convection_ra1e4_results},
\ref{tab:convection_ra1e5_results}, and
\ref{tab:convection_ra1e6_results} give the results of these
computations. It should be noted that the negative pressures computed
are due to a relative pressure being reported. For each of the data
points presented, both Newton-Krylov solutions and FANM solutions
computed identical numerical results for all of these global
values. No variation in solution was noted for each of the
computations larger than the tolerance of the converged nonlinear
residual norms given in
Table~\ref{tab:convection_residual_norm_comparison}. Variations in the
convergence behavior and differences in performance for these
solutions will be analyzed in the following section.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and a FANM solver for Reynolds numbers
of 100, 300, 500, and 700 on a $50 \times 50$ square
grid. Table~\ref{tab:thermal_convection_parameters} gives the physical
parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Reynolds
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} fixed and varying the
fluid velocity in the $x$ direction on the top boundary. In addition
to multigrid preconditioning for the linear solution at each Newton
step, a polynomial line search method was used as a means of
globalization assisting the Newton correction computation as outlined
by Pawlowski in \citep{pawlowski_globalization_2006} while the forcing
term was again computed using method 2 from Shadid's work
\citep{shadid_inexact_1997}.

@@

In the literature on this benchmark, solutions at Reynolds numbers of
up to \sn{1}{4} were achieved using a variety of nonlinear solution
techniques. With the FANM method, at Reynolds numbers above 700
convergence of the linear iterations could not be achieved regardless
of how agressive the multigrid preconditioner was set. In these cases,
the Jacobian operators are ill-conditioned enough that even GMRES
converges slowly relative to the thermal convection cavity
benchmark. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case and others where transport driven
by inertial forces is dominant. Because of this, only solutions to
the lid driven cavity benchmark at Reynolds numbers of up to 700 are
presented.

@@

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
method. For each benchmark, we will analyze the iterative performance
of both the Newton solver and the linear solver used to compute the
correction vector at each step. In addition, timing results will be
discussed.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method. Limiting the number of nonlinear iterations required for
solution is critical for efficiency as nonlinear function evaluations,
Jacobian generation, and preconditioner construction are typically
required at every Newton iteration. For all Rayleigh numbers except
\sn{1}{5} the number of Newton iterations required to converge is the
same while the \sn{1}{5} result required an extra FANM iteration. This
is due to the fact that the convergence of the linear residual has a
strong effect on the convergence of the nonlinear residual, much
stronger than the effect on the convergence of the eigenvalue in the
$SP_N$ computations where we observed that identical numbers of
eigenvalue iterations were required to converge each variation of the
probem.

As outlined in our discussion on inexact Newton methods, depending on
how converged the linear residual is at each step, it is possible to
throw off the nonlinear iteration and decrease iterative
performance. Although the same preconditioner was used with both the
Newton-Krylov and FANM calculations, the convergence properties of
MCSA and GMRES are different and we therefore expect them to produce
different results at each nonlinear iteration which will in turn
affect the forcing term calculation. If they are different enough, it
is possible that another Newton iteration may be required. Even
considering the extra Newton iteration, the solutions are converged to
effectively the same residual norm tolerance and produce the same
results as provided by the verification.

Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
behavior for the \sn{1}{5} Rayleigh number case that required the
extra FANM iteration. We see that the extra iteration comes from
slightly slower quadratic convergence with FANM for this particular
case. The difference in convergence of the nonlinear problem comes
from the fact that the same multigrid preconditioner parameters are
used for this case as for the \sn{1}{3} and \sn{1}{4} Rayleigh number
cases for the linear solutions. Using the same preconditioner for the
harder \sn{1}{5} case means that the problem is still growing in
difficulty even though it is preconditioned. The more ill-conditioned
the linear model is, the closer the linear residual is converged to
the requested tolerance whereas a better conditioned problem may
produce a linear solution that overshoots the specified tolerance
provided by the forcing term. When oversolving is not an issue, this
results in faster convergence of the nonlinear problem.

For the \sn{1}{5} Rayleigh number case, this means that the GMRES
solver at each Newton step is converging the linear residual to a
smaller tolerance than the MCSA solver, even though they have
approximately the same forcing terms to begin with. As the Rayleigh
number increases, the problem becomes more difficult to solve and
therefore a more agressive preconditioner should maintain iterative
performance of the nonlinear method. This was observed to be true for
the \sn{1}{6} Rayleigh number case. For this calculation, the number
of multigrid cycle applications was bumped up to 3 from 1 and FANM
converged in the same number of nonlinear iterations as the
Newton-Krylov method.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear
iterations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solvers preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme. Even the
\sn{1}{5} Rayleigh number case that required more Newton iterations
required 5 fewer MCSA iterations to converge than GMRES iterations.

@@

Although the iterative performance of FANM is excellent for the
thermal convection cavity problem, the explicit preconditioning
strategy causes the same issues as was observed for the $SP_N$
equations in the previous
chapter. Table~\ref{tab:convection_speedup_comparison} gives the
speedup of the Newton-Krylov method over FANM. For all Rayleigh
numbers, we see a fairly constant speedup of O(400) when using
Newton-Krylov over FANM meaning that the implementation of the FANM
method used here is significantly slower. As the inexact Newton scheme
in Drekar was modified by simply swapping out the linear solver used
at each Newton step, this increased compute time is coming directly
from using MCSA as the solver and the overhead due to
preconditioning. When preconditioning overhead for the $SP_N$
equations was considered as shown by the data in
Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
approximately two orders of magnitude was noted between the Krylov
methods and MCSA showing consistency between those results and the
fluid benchmark results. In addition, the fairly constant speedup
values as a function of the primary problem parameter show that FANM
has effectively the same time complexity as a Newton-Krylov method and
the implementation timings simply differ by a constant value.

@@

For the lid driven cavity benchmark, performance results were again
compared between the FANM and Newton-Krylov calculations. In this
case, the nonlinear iterative performance results as given by
Table~\ref{tab:driven_nonlinear_iter_comparison} show that the same
number of Newton iterations were required to converge the 100, 300,
and 500 Reynolds number cases while the FANM calculation with a
Reynolds number of 700 actually converged in 4 fewer iterations. The
nonlinear residual norm as a function of Newton iteration for this
case is given in Figure~\ref{fig:re700_convergence}.

@@

The difficulty of this benchmark problem as compared to the thermal
convection cavity problem is readily apparent in the linear iteration
counts as given by Table~\ref{tab:driven_linear_iter_comparison}. As
the opposite of what was observed for that benchmark, lid driven
cavity results show that the linear models were significantly more
ill-conditioned, requiring many more MCSA and GMRES iterations. For
all cases except the 700 Reynolds number case, many more MCSA
iterations were required to converge with the observed GMRES iteration
count 64\% of that observed for MCSA when looking at the 100 Reynolds
number case. For the 700 Reynolds number case, fewer MCSA iterations
were observed due to the fact that 4 fewer nonlinear iterations were
performed. The average number of linear iterations per nonlinear
iteration was still much larger for the FANM calculations.

These iterative results again point out the fundamental flaw of
requiring a spectral radius of less than unity for the MCSA iteration
to converge at each Newton step. As the problem becomes more dominated
by inertial forces, the resulting linear models become increasingly
stiff and difficult to precondition. As a result, convergence requires
many more iterations than needed for the flow problem driven by
convection in the previous benchmark if convergence is achieved at
all.
